Reanalysis Step 2: Multiple Imputation (Clean Version)

### 2025-08-29 reanalysis-2-multiple_imputation.qmd
more predictor

---
title: "Reanalysis Step 2: Multiple Imputation"
subtitle: "Imputing missing values using mice package"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

```{r setup}
#| include: false
source("dependencies.R")
source("functions.R")  # Load formula definitions and helper functions
seed <- 2025
setup_analysis(seed = seed)

# Additional packages for imputation
library(mice)
library(miceadds)  # For additional 2-level imputation methods
library(VIM)
library(naniar)

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 2 iterations (fast)
# final = 30 imputations, 20 iterations (stability achieved ~20)
analysis_mode <- "test"  # Change to "final" for production run

# Set parameters based on mode
if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 2
} else if (analysis_mode == "final") {
  n_imputations <- 30
  n_iterations <- 20
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}
```

**Running in `r toupper(analysis_mode)` mode:**  
- **Imputations:** `r n_imputations`  
- **Iterations:** `r n_iterations`

## Load Prepared Data

```{r load-data}
# Load the prepared dataset from Step 1
df_for_imputation <- readRDS(file.path(reanalysis_data_dir, "df_for_imputation.rds"))
variable_lists <- readRDS(file.path(reanalysis_data_dir, "variable_lists.rds"))

ps_model_vars <- variable_lists$ps_model_vars
outcome_model_vars <- variable_lists$outcome_model_vars
```

## Create Fellow Eye Variables

```{r create-fellow-eye-vars}
# Identify patients with both eyes and create fellow eye predictors
# This leverages the correlation between eyes from the same patient
# to improve imputation plausibility

df_with_fellow <- df_for_imputation |>
  group_by(e_mrn_deidentified) |>
  mutate(
    n_eyes = n(),
    has_fellow = n_eyes == 2,
    eye_index = row_number(),

    # Get fellow eye values (NA if patient has only 1 eye)
    # Using [3 - eye_index] to get the other eye (works for 2 eyes: eye 1 gets eye 2, eye 2 gets eye 1)
    fellow_baseline_VA = if_else(n_eyes == 2, 
                                 baseline_VA_logMAR[3 - eye_index], 
                                 NA_real_),
    fellow_intermediate_VA = if_else(n_eyes == 2,
                                    intermediate_VA_logMAR[3 - eye_index],
                                    NA_real_),
    fellow_outcome_VA = if_else(n_eyes == 2,
                               outcome_VA_logMAR[3 - eye_index],
                               NA_real_),
    fellow_lapsed = if_else(n_eyes == 2,
                           ever_lapse_binary[3 - eye_index],
                           NA_real_),
    # Note: fellow_dr needs special handling to maintain factor structure
    fellow_dr_numeric = if_else(n_eyes == 2,
                               person_dr[3 - eye_index],
                               NA_real_),

    # Patient-level summaries (same for both eyes of a patient)
    patient_worst_baseline = max(baseline_VA_logMAR, na.rm = TRUE),
    patient_best_baseline = min(baseline_VA_logMAR, na.rm = TRUE),
    patient_ever_lapsed = max(ever_lapse_binary, na.rm = TRUE)
  ) |>
  ungroup() |>
  mutate(
    # Handle Inf values from max/min with all NAs
    patient_worst_baseline = if_else(is.infinite(patient_worst_baseline), NA_real_, patient_worst_baseline),
    patient_best_baseline = if_else(is.infinite(patient_best_baseline), NA_real_, patient_best_baseline),

    # Convert fellow_dr_numeric to ordered factor matching person_dr structure
    fellow_dr = factor(fellow_dr_numeric, levels = c(0, 1, 2),
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE)
  ) |>
  select(-fellow_dr_numeric)  # Remove the temporary numeric version

# Summary of eye pairing
n_patients <- n_distinct(df_with_fellow$e_mrn_deidentified)
n_single_eyes <- sum(df_with_fellow$n_eyes == 1)
n_paired_eyes <- sum(df_with_fellow$n_eyes == 2)

# Log patient/eye summary
log_message("Patient/Eye Summary:")
log_message(sprintf("- Total patients: %d", n_patients))
log_message(sprintf("- Patients with 1 eye: %d (%.1f%%)", 
            n_single_eyes, 100 * n_single_eyes / nrow(df_with_fellow)))
log_message(sprintf("- Patients with 2 eyes: %d (%.1f%%)", 
            n_paired_eyes/2, 100 * n_paired_eyes / nrow(df_with_fellow)))
log_message(sprintf("- Total eyes: %d", nrow(df_with_fellow)))

# Display summary for output
patient_eye_summary <- tibble(
  Category = c("Total patients", "Patients with 1 eye", "Patients with 2 eyes", "Total eyes"),
  N = c(n_patients, n_single_eyes, n_paired_eyes/2, nrow(df_with_fellow)),
  Percent = c(NA, 100 * n_single_eyes / nrow(df_with_fellow), 
              100 * n_paired_eyes / nrow(df_with_fellow), NA)
)
kable(patient_eye_summary, caption = "Patient/Eye Distribution", digits = 1)

# Check fellow eye data availability
fellow_summary <- df_with_fellow |>
  filter(n_eyes == 2) |>
  summarise(
    n_fellow_baseline = sum(!is.na(fellow_baseline_VA)),
    n_fellow_intermediate = sum(!is.na(fellow_intermediate_VA)),
    n_fellow_outcome = sum(!is.na(fellow_outcome_VA)),
    pct_fellow_baseline = round(100 * n_fellow_baseline / n(), 1),
    pct_fellow_intermediate = round(100 * n_fellow_intermediate / n(), 1),
    pct_fellow_outcome = round(100 * n_fellow_outcome / n(), 1)
  )

# Log fellow eye data availability
log_message("Among paired eyes, fellow eye data availability:")
log_message(sprintf("- Fellow baseline VA: %.1f%%", fellow_summary$pct_fellow_baseline))
log_message(sprintf("- Fellow intermediate VA: %.1f%%", fellow_summary$pct_fellow_intermediate))
log_message(sprintf("- Fellow outcome VA: %.1f%%", fellow_summary$pct_fellow_outcome))

# Display as table
fellow_availability <- tibble(
  Measure = c("Fellow baseline VA", "Fellow intermediate VA", "Fellow outcome VA"),
  Percent_Available = c(fellow_summary$pct_fellow_baseline, 
                        fellow_summary$pct_fellow_intermediate, 
                        fellow_summary$pct_fellow_outcome)
)
kable(fellow_availability, caption = "Fellow Eye Data Availability (Among Paired Eyes)", digits = 1)

# Use this enhanced dataset for imputation
df_for_imputation <- df_with_fellow
```

@@

```{r define-imputation-vars}
# Variables to include in imputation model
# Include: outcome, exposure, all covariates source variables only
# Derived variables will be recreated post-imputation

# Core analysis variables - SOURCE ONLY
imputation_vars <- c(
  "e_mrn_deidentified",

  # Temporal VA variables
  "cohort_id",  # Data indicator only - not used as predictor
  "baseline_VA_logMAR",  # Baseline VA
  "intermediate_VA_logMAR",  # Intermediate follow-up VA (for excluded cohort)
  "outcome_VA_logMAR",  # 2-year outcome (true for submitted, missing for excluded)
  # NOT outcome_va_vi_binary - derived from logMAR >= 0.3

  # Exposure
  "ever_lapse_binary",

  # Demographics/Clinical
  "gender_cat",
  "race_ethnic_cat",
  "insurance_cat",
  "age",  # Continuous age (source)
  # NOT age_cat - derived from cut(age)
  "CCI",
  "DCSI",

  # Treatment indicators - ALL SOURCES
  "other_inject",
  "anti_VEGF",
  "focal_laser_flag",
  "PRP_flag",
  # NOT treatment_type or any_treatment - derived from these 4

  # Comorbidities (already recoded to 0/1)
  "glaucoma_bef_hitplus_cat",
  "otherretina_bef_hitplus_cat",
  "catsurg_before_hitplus_cat",

  # DR severity - use person_dr (created from No_DR/NPDR/PDR)
  "person_dr",  # Numeric 0/1/2
  # NOT dr_severity or person_dr_severity - derived from person_dr

  # Fellow eye predictors (if available)
  "fellow_baseline_VA",
  "fellow_intermediate_VA",
  "fellow_outcome_VA",
  # "fellow_lapsed",  # REMOVED: Collinear with ever_lapse_binary
  "fellow_dr",
  "n_eyes",
  # "has_fellow",  # REMOVED: Perfectly collinear with n_eyes (has_fellow = n_eyes == 2)
  "patient_worst_baseline",
  "patient_best_baseline"
  # "patient_ever_lapsed"  # REMOVED: Collinear with ever_lapse_binary
)

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    cohort_id = factor(cohort_id),  # Ensure cohort_id is a factor
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  )
```

## Setup Imputation Model

```{r setup-mice}
# Initialize mice
# First, do a dry run to check the imputation setup
mice_dry <- mice(df_mice, maxit = 0, printFlag = FALSE)

# Imputation model summary
miss_vars <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  Pct_Missing = round(mice_dry$nmis/nrow(df_mice) * 100, 1)
) |>
  filter(N_Missing > 0) |>
  arrange(desc(N_Missing))
```

### Initial Imputation Model Summary

**Total variables in dataset:** `r ncol(df_mice)`  
**Variables to be imputed:** `r sum(mice_dry$nmis > 0)`  
**Total missing values:** `r sum(mice_dry$nmis)`

**Missing data by variable:**
```{r missing-vars-table}
kable(miss_vars)

method_table <- data.frame(
  Variable = names(mice_dry$method),
  Method = mice_dry$method,
  stringsAsFactors = FALSE
) |>
  filter(Method != "")
```

**Initial imputation methods by variable:**
```{r method-table}
kable(method_table)

# Customize imputation methods as needed
custom_methods <- mice_dry$method

# Debug: Check if any methods have invalid values
# Include all standard MICE methods plus 2-level methods from mice and miceadds
valid_methods <- c("", "pmm", "logreg", "polyreg", "polr", "norm", "norm.boot", "norm.nob", 
                   "norm.predict", "sample", "cart", "rf", 
                   # 2-level methods
                   "2l.norm", "2l.lmer", "2l.pan", "2lonly.mean", "2lonly.norm", "2lonly.pmm",
                   # miceadds methods
                   "2l.pmm", "2l.2stage.norm", "2l.2stage.pmm", "ml.lmer")
invalid_methods <- custom_methods[!custom_methods %in% valid_methods]
if (length(invalid_methods) > 0) {
  log_message(paste0("WARNING: Invalid imputation methods detected: ", paste(names(invalid_methods), "=", invalid_methods, collapse=", ")))
  # Reset any invalid methods to appropriate defaults
  for (var in names(invalid_methods)) {
    if (var %in% names(custom_methods)) {
      # Set to empty string to skip imputation for now
      custom_methods[var] <- ""
      log_message(paste0("Reset method for ", var, " to '' (no imputation)"))
    }
  }
}

# Ensure methods vector has proper names
if (is.null(names(custom_methods))) {
  log_message("ERROR: custom_methods has no names. Setting names from df_mice columns.")
  names(custom_methods) <- colnames(df_mice)
}

# Print diagnostic info
log_message(paste0("Number of variables in custom_methods: ", length(custom_methods)))
log_message(paste0("First few method assignments: ", 
                  paste(head(names(custom_methods)), "=", head(custom_methods), collapse=", ")))

# Temporal VA variables - use 2-level PMM for clustered data
# 2l.pmm from miceadds: Multilevel PMM using linear mixed models
# This properly accounts for correlation between eyes from same patient
custom_methods["baseline_VA_logMAR"] <- "2l.pmm"  # Multilevel PMM for eyes within patients
custom_methods["intermediate_VA_logMAR"] <- "2l.pmm"  # Multilevel PMM for eyes within patients
custom_methods["outcome_VA_logMAR"] <- "2l.pmm"  # Multilevel PMM for eyes within patients
custom_methods["cohort_id"] <- ""  # Never impute - data management indicator only

# Other continuous variables
# CCI and DCSI could potentially vary by eye visit, so use 2-level
custom_methods["CCI"] <- "2l.pmm"  # Multilevel PMM for clustered data
custom_methods["DCSI"] <- "2l.pmm"  # Multilevel PMM for clustered data
custom_methods["age"] <- "pmm"  # Regular PMM (patient-level, doesn't vary by eye)

# Use proportional odds model for ordinal person_dr (0/1/2)
custom_methods["person_dr"] <- "polr"  # Ordinal logistic regression

# Use logistic regression for binary variables
custom_methods["catsurg_before_hitplus_cat"] <- "logreg"

# Use polytomous regression for categorical variables with >2 levels
custom_methods["race_ethnic_cat"] <- "polyreg"
custom_methods["insurance_cat"] <- "polyreg"

# Don't impute these variables
# Method = "" means: don't impute (but still use as predictor EXCEPT for IDs)
custom_methods["row_id"] <- ""  # Row ID - never impute, never use as predictor; see below.
custom_methods["e_mrn_deidentified"] <- ""  # Patient ID - never impute, never use as predictor; see below.

# Fellow eye variables should not be imputed (they are derived from existing data)
# NOTE: fellow_outcome_VA may show collinearity warning with outcome_VA_logMAR
# This is expected - they have identical missingness patterns (both NA for excluded cohort)
custom_methods["fellow_outcome_VA"] <- ""
custom_methods["fellow_baseline_VA"] <- ""
custom_methods["fellow_intermediate_VA"] <- ""
# custom_methods["fellow_lapsed"] <- ""  # REMOVED - collinear with lapse status
custom_methods["fellow_dr"] <- ""
custom_methods["n_eyes"] <- ""
# custom_methods["has_fellow"] <- ""  # REMOVED - collinear with n_eyes
custom_methods["patient_worst_baseline"] <- ""
custom_methods["patient_best_baseline"] <- ""
# custom_methods["patient_ever_lapsed"] <- ""  # REMOVED - collinear with ever_lapse_binary

# Variables that are complete (no missing data) - set method but won't actually be imputed
# Setting methods ensures correct measurement level is used IF they had missing data
custom_methods["ever_lapse_binary"] <- "logreg"  # Binary - complete
custom_methods["glaucoma_bef_hitplus_cat"] <- "logreg"  # Binary - already recoded to 0/1
custom_methods["otherretina_bef_hitplus_cat"] <- "logreg"  # Binary - already recoded to 0/1
custom_methods["gender_cat"] <- "logreg"  # Binary categorical - likely complete
custom_methods["other_inject"] <- "logreg"  # Binary - likely complete
custom_methods["anti_VEGF"] <- "logreg"  # Binary - likely complete  
custom_methods["focal_laser_flag"] <- "logreg"  # Binary - likely complete
custom_methods["PRP_flag"] <- "logreg"  # Binary - likely complete

# Note: Variables with complete data won't be imputed even if they have a method assigned
# The method specification ensures proper measurement level if imputation were needed

# Check predictor matrix (which variables predict which)
pred_matrix <- mice_dry$predictorMatrix

# Mark e_mrn_deidentified as cluster variable
# -2 indicates this is the cluster ID (patient ID) for 2-level imputation
# This is REQUIRED for 2l.pmm to work properly
# The 2l.pmm method from miceadds:
# - Uses linear mixed models (lmer) to model the multilevel structure
# - Preserves within-patient correlation between eyes
# - Obtains PMM donors across all clusters by default
# - Properly accounts for the hierarchical data structure
pred_matrix[, "e_mrn_deidentified"] <- -2
pred_matrix["e_mrn_deidentified", ] <- 0  # Don't predict the cluster ID itself
pred_matrix[, "row_id"] <- 0
pred_matrix["row_id", ] <- 0

# TEMPORAL STRUCTURE RULES
# 1. Baseline can predict intermediate and outcome
pred_matrix["intermediate_VA_logMAR", "baseline_VA_logMAR"] <- 1
pred_matrix["outcome_VA_logMAR", "baseline_VA_logMAR"] <- 1

# 2. Intermediate can predict outcome
# NOTE: MICE predictor matrix only uses 0/1, not weights.
# To prioritize intermediate VA over baseline for outcome prediction:
# - Both are included as predictors (temporal proximity naturally helps)
# - PMM will naturally weight closer timepoints more heavily
# - Alternative: Could create interaction terms or use method="norm.boot" with custom weights
pred_matrix["outcome_VA_logMAR", "intermediate_VA_logMAR"] <- 1

# 3. CRITICAL: Prevent reverse-time prediction
pred_matrix["intermediate_VA_logMAR", "outcome_VA_logMAR"] <- 0  # Outcome cannot predict intermediate
pred_matrix["baseline_VA_logMAR", "outcome_VA_logMAR"] <- 0  # Outcome cannot predict baseline
pred_matrix["baseline_VA_logMAR", "intermediate_VA_logMAR"] <- 0  # Intermediate cannot predict baseline

# Cohort_id should not be predicted by anything AND should not predict anything
pred_matrix["cohort_id", ] <- 0  # Not predicted
pred_matrix[, "cohort_id"] <- 0  # Not a predictor

# Don't use variables as their own predictors
diag(pred_matrix) <- 0

# Ensure exposure is used as predictor for outcomes
pred_matrix[, "ever_lapse_binary"] <- 1
pred_matrix["ever_lapse_binary", "ever_lapse_binary"] <- 0

# DR severity predicts visual acuity at all timepoints
if ("person_dr" %in% colnames(pred_matrix)) {
  pred_matrix["baseline_VA_logMAR", "person_dr"] <- 1
  pred_matrix["intermediate_VA_logMAR", "person_dr"] <- 1
  pred_matrix["outcome_VA_logMAR", "person_dr"] <- 1
}

# FELLOW EYE STRUCTURE
# Fellow eye outcomes strongly predict index eye outcomes (correlation typically 0.7-0.9)
# NOTE: MICE predictor matrix only uses 0/1, not weights. 1 = used as predictor, 0 = not used
if ("fellow_outcome_VA" %in% colnames(pred_matrix)) {
  # Fellow eye visual acuity predicts same timepoint
  pred_matrix["outcome_VA_logMAR", "fellow_outcome_VA"] <- 1
  pred_matrix["baseline_VA_logMAR", "fellow_baseline_VA"] <- 1  
  pred_matrix["intermediate_VA_logMAR", "fellow_intermediate_VA"] <- 1

  # Fellow baseline also helps predict later timepoints
  pred_matrix["outcome_VA_logMAR", "fellow_baseline_VA"] <- 1
  pred_matrix["intermediate_VA_logMAR", "fellow_baseline_VA"] <- 1

  # DR severity correlation between eyes
  if ("person_dr" %in% rownames(pred_matrix)) {
    pred_matrix["person_dr", "fellow_dr"] <- 1
  }

  # Don't let fellow eye variables be predicted by anything
  # Only set for variables that actually exist in the matrix
  fellow_vars_in_matrix <- c("fellow_outcome_VA", "fellow_baseline_VA", "fellow_intermediate_VA", "fellow_dr")
  fellow_vars_in_matrix <- fellow_vars_in_matrix[fellow_vars_in_matrix %in% rownames(pred_matrix)]
  if (length(fellow_vars_in_matrix) > 0) {
    pred_matrix[fellow_vars_in_matrix, ] <- 0
  }

  # Don't let fellow eye variables be their own predictors
  if ("fellow_outcome_VA" %in% rownames(pred_matrix)) {
    pred_matrix["fellow_outcome_VA", "fellow_outcome_VA"] <- 0
  }
  if ("fellow_baseline_VA" %in% rownames(pred_matrix)) {
    pred_matrix["fellow_baseline_VA", "fellow_baseline_VA"] <- 0
  }
  if ("fellow_intermediate_VA" %in% rownames(pred_matrix)) {
    pred_matrix["fellow_intermediate_VA", "fellow_intermediate_VA"] <- 0
  }
}

# Patient-level variables can predict eye-level outcomes (but not strongly)
if ("patient_worst_baseline" %in% colnames(pred_matrix)) {
  pred_matrix[, "patient_worst_baseline"] <- 1
  pred_matrix[, "patient_best_baseline"] <- 1
  # pred_matrix[, "patient_ever_lapsed"] <- 1  # REMOVED - variable not in imputation

  # But don't predict these variables
  pred_matrix["patient_worst_baseline", ] <- 0
  pred_matrix["patient_best_baseline", ] <- 0
  # pred_matrix["patient_ever_lapsed", ] <- 0  # REMOVED - variable not in imputation
}

# n_eyes is just an indicator, not a predictor
if ("n_eyes" %in% colnames(pred_matrix)) {
  pred_matrix[, "n_eyes"] <- 0
  pred_matrix["n_eyes", ] <- 0
  # has_fellow removed from imputation - was collinear with n_eyes
}

```

### Final Imputation Model Configuration

```{r final-model-config}

# Get actual missing data information
missing_info <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  stringsAsFactors = FALSE
)

>>>>

---
title: "Reanalysis Step 2: Multiple Imputation (Clean Version)"
subtitle: "Simplified imputation using mice with clustering"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

```{r setup}
#| include: false
source("dependencies.R")
source("functions.R")
seed <- 2025
setup_analysis(seed = seed)

library(mice)
library(miceadds)
library(VIM)
library(naniar)
library(dplyr)
library(tidyr)
library(knitr)
library(lme4)

# Initialize logging
log_file <- init_log("multiple_imputation_clean")

# ANALYSIS MODE
analysis_mode <- "final"  # "test" or "final"

if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 3
} else {
  n_imputations <- 30
  n_iterations <- 20
}

log_message(paste0("Analysis mode: ", analysis_mode))
log_message(paste0("Imputations: ", n_imputations, ", Iterations: ", n_iterations))
```

## Load Data

```{r load-data}
# Load the prepared dataset
df_original <- readRDS(file.path(reanalysis_data_dir, "df_for_imputation.rds"))
log_message(paste0("Loaded ", nrow(df_original), " observations"))

# Get unique patient count
n_patients <- n_distinct(df_original$e_mrn_deidentified)
log_message(paste0("Number of unique patients: ", n_patients))
```

@@

```{r define-variables}
# Variables that should NOT be imputed and should NOT be used as predictors
vars_no_impute_no_predict <- c(
  "row_id"  # Internal row identifier
)

# Variables that should NOT be imputed but CAN be used as predictors  
vars_no_impute_yes_predict <- c(
  "ever_lapse_binary",  # Exposure - complete
  "cohort_id"          # Data indicator - complete
)

# Cluster variable (patient ID)
cluster_var <- "e_mrn_deidentified"

# Variables to impute - continuous (use PMM with clustering)
vars_impute_continuous <- c(
  "baseline_VA_logMAR",
  "intermediate_VA_logMAR", 
  "outcome_VA_logMAR",
  "age",
  "CCI",
  "DCSI"
)

# Variables to impute - binary
vars_impute_binary <- c(
  "catsurg_before_hitplus_cat"
)

# Variables to impute - categorical
vars_impute_categorical <- c(
  "gender_cat",
  "race_ethnic_cat",
  "insurance_cat"
)

# Variables to impute - ordinal
vars_impute_ordinal <- c(
  "person_dr"  # 0=No_DR, 1=NPDR, 2=PDR
)

# Treatment variables (binary, likely complete)
vars_treatment <- c(
  "other_inject",
  "anti_VEGF",
  "focal_laser_flag",
  "PRP_flag"
)

# Comorbidity indicators (binary factors from step 1)
vars_comorbidity <- c(
  "glaucoma_bef_hitplus_cat",
  "otherretina_bef_hitplus_cat"
)

# All variables for imputation dataset
all_vars <- c(
  cluster_var,
  vars_no_impute_no_predict,
  vars_no_impute_yes_predict,
  vars_impute_continuous,
  vars_impute_binary,
  vars_impute_categorical,
  vars_impute_ordinal,
  vars_treatment,
  vars_comorbidity
)
```

## Prepare Data for MICE

```{r prepare-data}
# Select variables and clean up (rely on step 1 outputs)
df_mice <- df_original |>
  select(any_of(all_vars)) |>
  mutate(
    # Ensure person_dr is ordered factor
    person_dr = if (is.numeric(person_dr)) {
      factor(person_dr, levels = c(0L, 1L, 2L), labels = c("No_DR", "NPDR", "PDR"), ordered = TRUE)
    } else {
      factor(person_dr, levels = c("No_DR", "NPDR", "PDR"), ordered = TRUE)
    },
    # Clean factor levels (remove spaces for formula compatibility)
    across(where(is.character), ~ factor(gsub("[^[:alnum:]_]", "_", .x)))
  ) |>
  dplyr::group_by(e_mrn_deidentified) |>
  dplyr::mutate(
    n_eyes = dplyr::n(),
    eye_index = dplyr::row_number(),
    fellow_baseline_VA = dplyr::if_else(n_eyes == 2, baseline_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_intermediate_VA = dplyr::if_else(n_eyes == 2, intermediate_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_outcome_VA = dplyr::if_else(n_eyes == 2, outcome_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_dr = dplyr::if_else(n_eyes == 2,
                               person_dr[3 - eye_index],
                               factor(NA, levels = levels(person_dr), ordered = TRUE))
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-eye_index) |>
  dplyr::mutate(
    # Convert cluster variable to integer for 2l methods
    e_mrn_deidentified = as.integer(as.factor(e_mrn_deidentified))
  )

# Check missing data pattern
log_message("Missing data summary:")
missing_summary <- df_mice |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))

print(missing_summary)
```

## Setup Imputation Methods

```{r setup-methods}
# Initialize mice to get default setup
ini <- mice(df_mice, maxit = 0, printFlag = FALSE)

# Set up methods vector
methods <- ini$method

# No imputation for these
methods[intersect(vars_no_impute_no_predict, names(methods))] <- ""
methods[intersect(vars_no_impute_yes_predict, names(methods))] <- ""
if (cluster_var %in% names(methods)) {
  methods[cluster_var] <- ""
}

# Continuous variables - use standard PMM (no multilevel)
for (var in vars_impute_continuous) {
  if (var %in% names(methods)) {
    methods[var] <- "pmm"
  }
}

# Binary variables
for (var in vars_impute_binary) {
  if (var %in% names(methods)) {
    methods[var] <- "logreg"
  }
}

# Categorical variables  
for (var in vars_impute_categorical) {
  if (var %in% names(methods)) {
    methods[var] <- "polyreg"
  }
}

# Ordinal variables
for (var in vars_impute_ordinal) {
  if (var %in% names(methods)) {
    methods[var] <- "polr"
  }
}

# Treatment and comorbidity variables (likely complete, but set methods just in case)
for (var in c(vars_treatment, vars_comorbidity)) {
  if (var %in% names(methods)) {
    methods[var] <- "logreg"
  }
}

# Fellow-eye variables: predictors only (do not impute)
vars_fellow <- c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")
for (var in vars_fellow) {
  if (var %in% names(methods)) {
    methods[var] <- ""
  }
}

log_message("Imputation methods configured")
```

## Setup Predictor Matrix

```{r setup-predictors}
# Use quickpred to get smart predictor selection
# This selects predictors with correlation >= 0.10 (R^2 >= 0.01)
# As recommended in Van Buuren's MICE documentation

# Variables that should always be included as predictors (key variables)
vars_always_predict <- c(
  "ever_lapse_binary",      # Exposure
  "baseline_VA_logMAR",     # Key outcome
  "outcome_VA_logMAR",      # Key outcome
  "age",                    # Demographics
  "gender_cat",
  "race_ethnic_cat",        # Demographics
  "insurance_cat",          # Demographics/SES
  "CCI", "DCSI",           # Comorbidities
  "person_dr",              # DR severity - key clinical variable
  "anti_VEGF",              # Key treatment
  "focal_laser_flag",       # Treatment
  "PRP_flag"                # Treatment
)

# Variables that should never be predictors
vars_never_predict <- c(
  vars_no_impute_no_predict
)

# Generate predictor matrix using quickpred
# Lower threshold to 0.05 to capture more predictive relationships
# This includes variables with correlations >= 0.05 (R^2 >= 0.0025)
pred_matrix <- quickpred(
  df_mice, 
  exclude = vars_never_predict,
  include = vars_always_predict,
  mincor = 0.05,  # Lowered from 0.10 to include more predictors
  method = "pearson"
)

# No multilevel imputation: ensure cluster ID is not used as predictor
pred_matrix[, cluster_var] <- 0
pred_matrix[cluster_var, ] <- 0

# Ensure temporal ordering for VA variables
# Baseline can predict intermediate and outcome
pred_matrix["intermediate_VA_logMAR", "baseline_VA_logMAR"] <- 1
pred_matrix["outcome_VA_logMAR", "baseline_VA_logMAR"] <- 1

# Intermediate can predict outcome  
pred_matrix["outcome_VA_logMAR", "intermediate_VA_logMAR"] <- 1

# Prevent reverse-time prediction
pred_matrix["baseline_VA_logMAR", "outcome_VA_logMAR"] <- 0
pred_matrix["baseline_VA_logMAR", "intermediate_VA_logMAR"] <- 0
pred_matrix["intermediate_VA_logMAR", "outcome_VA_logMAR"] <- 0

# Add theoretically important relationships even if correlation is low
# Demographics should predict each other
demographic_vars <- c("age", "gender_cat", "race_ethnic_cat", "insurance_cat")
for (v1 in demographic_vars) {
  for (v2 in demographic_vars) {
    if (v1 != v2 && v1 %in% rownames(pred_matrix) && v2 %in% colnames(pred_matrix)) {
      pred_matrix[v1, v2] <- 1
    }
  }
}

# Comorbidities should predict DR severity and vice versa
if ("person_dr" %in% rownames(pred_matrix)) {
  if ("CCI" %in% colnames(pred_matrix)) pred_matrix["person_dr", "CCI"] <- 1
  if ("DCSI" %in% colnames(pred_matrix)) pred_matrix["person_dr", "DCSI"] <- 1
  if ("glaucoma_bef_hitplus_cat" %in% colnames(pred_matrix)) pred_matrix["person_dr", "glaucoma_bef_hitplus_cat"] <- 1
}

# Treatment variables should predict each other (treatment patterns)
treatment_vars <- c("anti_VEGF", "focal_laser_flag", "PRP_flag", "other_inject")
for (v1 in treatment_vars) {
  for (v2 in treatment_vars) {
    if (v1 != v2 && v1 %in% rownames(pred_matrix) && v2 %in% colnames(pred_matrix)) {
      pred_matrix[v1, v2] <- 1
    }
  }
}

# Fellow-eye predictors with weight=2 for stronger influence
fellow_vars <- c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr")
for (v in fellow_vars) {
  if (v %in% rownames(pred_matrix)) {
    pred_matrix[v, ] <- 0
  }
}
if ("fellow_baseline_VA" %in% colnames(pred_matrix)) {
  pred_matrix["baseline_VA_logMAR", "fellow_baseline_VA"] <- 2  # Weight=2
}
if ("fellow_intermediate_VA" %in% colnames(pred_matrix)) {
  pred_matrix["intermediate_VA_logMAR", "fellow_intermediate_VA"] <- 2  # Weight=2
}
if ("fellow_outcome_VA" %in% colnames(pred_matrix)) {
  pred_matrix["outcome_VA_logMAR", "fellow_outcome_VA"] <- 2  # Weight=2
}
if ("fellow_dr" %in% colnames(pred_matrix) && "person_dr" %in% rownames(pred_matrix)) {
  pred_matrix["person_dr", "fellow_dr"] <- 2  # Weight=2
}

log_message("Predictor matrix configured with temporal constraints")

# Print summary of predictor counts for each variable
pred_counts <- data.frame(
  Variable = rownames(pred_matrix),
  N_Predictors = rowSums(pred_matrix > 0),
  N_Predicting = colSums(pred_matrix > 0)[rownames(pred_matrix)]
)
pred_counts <- pred_counts[order(-pred_counts$N_Predictors), ]

log_message("\nTop variables by number of predictors:")
print(head(pred_counts, 15))
```

## Run Imputation

```{r run-imputation}
log_message("Starting MICE imputation...")

# Run the imputation
imp <- mice(
  df_mice, 
  m = n_imputations,
  method = methods,
  predictorMatrix = pred_matrix,
  maxit = n_iterations,
  seed = seed,
  printFlag = TRUE
)

log_message("MICE imputation completed")

# Check for logged events
if (!is.null(imp$loggedEvents)) {
  log_message("Logged events during imputation:")
  print(imp$loggedEvents)
}

# Save convergence plot
pdf(file.path(reanalysis_data_dir, "mice_convergence_clean.pdf"))
plot(imp)
dev.off()
```

## Extract and Save Results

```{r save-results}
# Extract all imputed datasets
imputed_list <- complete(imp, action = "all")

# Apply any derived variables needed
imputed_list <- lapply(imputed_list, function(data) {
  data |>
    mutate(
      # Create binary outcome from logMAR
      outcome_va_vi_binary = ifelse(outcome_VA_logMAR >= 1.0, 1, 0)
    )
})

# Save the mice object
saveRDS(imp, file.path(reanalysis_data_dir, "mice_results_clean.rds"))
log_message(paste0("Saved MICE object to: mice_results_clean.rds"))

# Save the list of complete datasets
saveRDS(imputed_list, file.path(reanalysis_data_dir, "imputed_datasets_clean.rds"))
log_message(paste0("Saved ", length(imputed_list), " imputed datasets"))

# Create summary
summary_stats <- data.frame(
  Metric = c(
    "Original observations",
    "Unique patients",
    "Variables in imputation", 
    "Number of imputations",
    "Iterations per imputation",
    "Variables with missing data",
    "Total missing values"
  ),
  Value = c(
    nrow(df_mice),
    n_patients,
    ncol(df_mice),
    n_imputations,
    n_iterations,
    sum(colSums(is.na(df_mice)) > 0),
    sum(is.na(df_mice))
  )
)

kable(summary_stats, caption = "Imputation Summary")

# Finalize logging
final_log <- finalize_log(success = TRUE)
```

**Analysis complete. Log file:** `r final_log`

## Next Steps

Run `reanalysis-3-imputation_diagnostics.qmd` to assess imputation quality.
